### 2022-06-10 paper.tex
adjust main section

We first discuss the task of quantifying the level of methylation in given, fixed, genomic interval. In Figure \ref{fig:smoothres}A, such an interval is covered by a single read for each of the two cells shown. The read from the cell 2 shows much more methylation than the read from cell 1, and a simple analysis would therefore consider that cell as having stronger methylation in the interval. However, given that the two reads agree wherever they overlap, a more parsimonious interpretation would be that the cells do not show difference in methylation within the interval. Rather, both cells, and similarly maybe most other cells, might have stronger methylation in the left third of the interval than in the middle one.

Therefore, we propose to not first obtain, for each CpG position, an average methylation across all cells and, then quantify for each individual cell its deviation from this average. In Figure \ref{fig:smoothres}B, the curved line depicts such an average over all cells, the red vertical lines show an individual cell's deviation from the ensemble average. We take the lengths of the red lines as signed values (``residual''), positive for lines extending upwards from the curve (methylated CpG) and negative for lines extending downwards (unmethylated CpG), and take the average over the residual for all the CpGs in the interval that are covered by reads from this cell. The value thus obtained is what we use to quantify this cell's (relative) methylation in the interval. For a genome tiled into such intervals, we thus obtain a matrix, one row per interval, one column per cell, that can be used for downstream analysis, e.g., as input for a PCA.

The signal to noise ratio in a matrix thus obtained will be much better than in a matrix obtained by simply averaging absolute methylation (0 or 1) over all the cell's covered CpG in a region. The reason for this is that we reduce the variation in situation as the one depicted in Figure \ref{fig:smoothres}A, where absolute methylation might differ strongly even though there is no actual evidence for a difference between the two cells.

@@

The standard PCA algorithm cannot deal with missing data. Therefore, one would need to make intervals wide enough to ensure that every interval is covered by at least one read for most windows and discard any interval that is not covered by any single cell. Hence, a naive analysis has to resort to a coarse tiling of the genome. To counter this, we propose a simple and straightforward way to deal with missing data in the input matrix to a PCA: In a first iteration, we replace each missing value in the centered input matrix with zero, then run the PCA. Then, these zeroes are replaced by the value predicted by the PCA and the PCA is rerun. Convergence is typically achieved after two to four iterations. For details, see Methods.

The possibility to perform PCA with missing data allows us to use a shorter and thus more intervals and hence provide the PCA with richer input data.

@@

Typically, some regions in a chromosome will have very similar methylation status in all cells (most commonly, fully methylated in nearly all cells) while other regions show variability in methylation across cells. Only the latter regions are of value for our goal of quantitating dissimilarity between cells. We call these the variably methylated regions (VMRs).

So far, we have discussed dividing up (tiling) each chromosome genome into non-overlapping, equal-sized intervals, and quantitating the methylation of each such window. Such rigid placing of interval boundaries is unlikely to be optimal: for example, a VMR might be much smaller than a tile and its signal will hence be drowned out by the larger part of uninformative CpGs that are equal in all cells, when averaging over all the CpG sites in the tile.

Therefore, we propose the following approach (Fig. \ref{vmr}): Divide up the chromosome into many \emph{overlapping} windows, that start at regular multiples of a fixed, small, step size. Quantify the methylation of each cell in each window by averaging the cell's methylation residuals over all CpGs in the window, as described above and depicted in Figure\ \ref{fig:smoothres}B. Then, calculate for each window the variance of these values over all cells. Select, say, the top XX\% windows with the highest variances and mark them as VMRs. Wherever thus marked windows overlap or are divided by only a small gap, merge them into one larger VMR. Then, calculate for each of these larger VMRs the methylation signal, as before, by averaging for each cell over the residuals of all contained CpGs.

In this manner, we obtain a methylation matrix, with one column per cell and one row per VMR, that is, in a sense, richer in information and has better signal-to-noise ratio than the matrix obtained by the simple analysis sketched at the very beginning. As we demonstrate below, a PCA performed on such a matrix provided a distance metric for the cells that contains more information on biological detail than one from a simpler analysis.

Besides providing better input for the distance calculations, the identified VMRs can also be compared with genomic annotation, thus providing a useful starting point for examining the data with methods from functional genomics.

\section{The scbs Python package}

>>>>

We first discuss the task of quantifying the level of methylation in a given, fixed, genomic interval. In Figure \ref{fig:smoothres}A, such an interval is covered by a single read for each of the two cells shown. The read from cell 2 shows much more methylation than the read from cell 1, and a simple analysis would therefore consider that cell as having stronger methylation in the interval. However, given that the two reads agree wherever they overlap, a more parsimonious interpretation would be that the cells do not show difference in methylation within the interval. Rather, both cells, and similarly maybe most other cells, might have stronger methylation in the left third of the interval than in the middle one.

Therefore, we propose to first obtain, for each CpG position, an average methylation across all cells and, then quantify for each individual cell its deviation from this average. In Figure \ref{fig:smoothres}B, the curved line depicts such an average over all cells, the red vertical lines show an individual cell's deviation from the ensemble average. We take the lengths of the red lines as signed values (``residual''), positive for lines extending upwards from the curve (methylated CpG) and negative for lines extending downwards (unmethylated CpG), and take the average over the residual for all the CpGs in the interval that are covered by reads from this cell. The value thus obtained is what we use to quantify this cell's (relative) methylation in the interval. For a genome tiled into such intervals, we thus obtain a matrix, one row per cell, one column per interval, that can be used for downstream analysis, e.g., as input for a PCA.

The signal to noise ratio in a matrix thus obtained will be much better than in a matrix obtained by simply averaging absolute methylation (0 or 1) over all the cell's covered CpG sites in a region. The reason for this is that we reduce the variation in situation as the one depicted in Figure \ref{fig:smoothres}A, where absolute methylation might differ strongly even though there is no actual evidence for a difference between the two cells.

@@

The standard PCA algorithm cannot deal with missing data. Therefore, one would need to make intervals wide enough to ensure that every interval is covered by at least one read for most windows and discard any interval that is not covered by any single cell. Hence, a naive analysis has to resort to a coarse tiling of the genome. To counter this, we propose a simple and straightforward way to deal with missing data in the input matrix to a PCA: In a first iteration, we replace each missing value in the centered input matrix with zero, then run the PCA. Then, only these zeroes are replaced by the value predicted by the PCA and the PCA is rerun. Convergence is typically achieved after two to five iterations. For details, see Methods.

The possibility to perform PCA with missing data allows us to use shorter, and thus more intervals and hence provide the PCA with richer input data.

@@

Typically, some regions in a chromosome will have very similar methylation status in all cells, while other regions show variability in methylation across cells. For instance, it is long known that CpG-rich promoters of housekeeping genes are unmethylated, and that a large proportion of the remaining genome is highly methylated regardless of cell type \citet{bird1986cpg}. In contrast, DNA methylation at certain genomic features such as enhancers is more dynamic \citep{argelaguet2019gastru}, and thus more variable across cells.
Only the latter regions are of value for our goal of quantitating dissimilarity between cells. We call these the variably methylated regions (VMRs).

So far, we have discussed dividing up (tiling) each chromosome into non-overlapping, equal-sized intervals, and quantitating the methylation of each such window. Such rigid placing of interval boundaries is unlikely to be optimal: for example, a VMR might be much smaller than a tile and its signal will hence be drowned out by the larger part of uninformative CpGs that are equal in all cells, when averaging over all the CpG sites in the tile.

Therefore, we propose the following approach (Fig. \ref{vmr}): Divide up the chromosome into many \emph{overlapping} windows, that start at regular multiples of a fixed, small, step size. Quantify the methylation of each cell in each window by averaging the cell's methylation residuals over all CpGs in the window, as described above and depicted in Figure\ \ref{fig:smoothres}B. Then, calculate for each window the variance of these values over all cells. Select, say, the top 2\% windows with the highest variances and mark them as VMRs. Wherever thus marked windows overlap or are divided by only a small gap, merge them into one larger VMR. Then, calculate for each of these larger VMRs the methylation signal, as before, by averaging for each cell over the residuals of all contained CpG sites.

In this manner, we obtain a methylation matrix, with one row per cell and one column per VMR, that is, in a sense, richer in information and has better signal-to-noise ratio than the matrix obtained by the simple analysis sketched at the very beginning. As we demonstrate below, a PCA performed on such a matrix provided a distance metric for the cells that contains more information on biological detail than one from a simpler analysis.

Besides providing better input for the distance calculations, the small size of identified VMRs also means that they can be readily compared with genomic annotation, which aids in their functional interpretation.
While a large 100~kb bin will often contain multiple genes with distinct expression patterns, small VMRs located in known regulatory elements or gene bodies are more easily associated with individual genes and their associated function.
Thus, VMRs provide a useful starting point for examining the data with methods from functional genomics.

\section{The \texttt{scbs} Python package}
